---
title: "R/tidyverse workshop part 2: ggplot2"
author: "Rose Driscoll"
date: "6/1/2022"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy = TRUE, tidy.opts=list(width.cutoff=60))

#install.packages("tidyverse")
library(tidyverse)
```

```{r}
# A few technical specifications:
# This workshop was designed and built using R 3.4.0, though it is expected to work with newer versions of R as well.
# It was tested with the following package versions:
#tidyverse 1.3.0
#cowplot 0.9.2
#gridGraphics 0.5-1
# See note line 698 about the behavior of cowplot versions <1.0.0 vs. newer versions
# RStudio (https://www.rstudio.com/) is recommended for viewing and running this Rmd file.
```


# ggplot2

**Topics**

1. The grammar of graphics (the philosophy behind ggplot2)
2. Aesthetics, mapping, geoms, etc (the nuts and bolts of ggplot2)
3. Making your plots pretty and readable
4. Exporting publication-ready, multipart figures


# PART 1: The grammar of graphics

`ggplot2` is based upon a layered grammar of graphics. A plot begins with some **data**, and is built in one or more layers. In each **layer**, variables from the data are **mapped** to different **aesthetics** of the plot, such as x, y, color, shape, etc. Each layer produces some sort of a **geometric object**, such as points, a line, bars, etc. In addition, the element(s) in each layer can be **positioned** in a particular way, and sometimes a **statistic** will be applied to create the object in the layer. Multiple layers in the same plot may all use the same data and mapping, or different data and mapping may be used in different layers. The plot also needs to have a **coordinate system**, and may optionally be **faceted** to display an additional categorical variable (we'll talk more about faceting later). 

For a much more detailed explanation of the layered grammar of graphics, see Wickham (2010).


# PART 2: Aesthetics, mapping, geoms, etc

The good news is that while the grammar of graphics gives you the power to control pretty much every aspect of your plot, you won't have to explicity specify all of those things in every single plot you make. ggplot2 provides handy defaults for almost everything, so the only things you are *required* to specify to make a plot are a dataset, a "geom" function to specify what type of geometric object should be created, and mappings of variables to plot aesthetics. We'll start with just these three things and work our way up.

Let's start with using the `iris` built-in dataset:

```{r}
data(iris)
head(iris)
```

Here's a simple example of ggplot2 syntax. This chunk has the option `eval = FALSE` set since we're not actually supplying data, a geom, or mapping yet - just marking out where those things will go. Don't try to run this code.

```{r, eval = FALSE}
ggplot(data = YOUR_DATA_HERE) +
  YOUR_GEOM_HERE(mapping = aes(YOUR_MAPPINGS_HERE))
```

Here's a first pass at making a scatterplot of sepal length versus sepal width using the `iris` data.

```{r}
ggplot(data = iris) +
  geom_point(mapping = aes(x = Sepal.Length, y = Sepal.Width))
```

The call to ggplot spans multiple lines, with the lines connected by a plus sign at the end of all but the last line. For now, the only argument we supply to the `ggplot()` function is the data frame. The second line is our geom function - for now, we are using `geom_point()` to make a scatterplot. We'll come back to different geom functions later.

## 1. Aesthetics and mapping

Let's start with aesthetics and mapping. In our sepal length vs. sepal width plot, we are mapping variables to two aesthetics: x and y.

```{r, eval = FALSE}
ggplot(data = iris) +
  geom_point(mapping = aes(x = Sepal.Length, y = Sepal.Width))
```

Mappings are supplied to the `mapping` argument, and need to be wrapped in the `aes()` function so that `ggplot` knows that they will become aesthetics of the plot. Inside the `aes()` function, we specify the aesthetic we want to map to followed by the name of the variable we want to map: `x = Sepal.Length`.

In our first example, we just used the `x` and `y` aesthetics, but there are lots of other aesthetics you can use, including `color`, `shape`, `size`, `alpha`, etc. The exact set of aesthetics available will depend on what geom you are using.

A few examples:

```{r}
ggplot(data = iris) +
  geom_point(mapping = aes(x = Sepal.Length, y = Sepal.Width, color = Species))

ggplot(data = iris) +
  geom_point(mapping = aes(x = Sepal.Length, y = Sepal.Width, shape = Species))
```

Some aesthetics are ordered, while others are unordered. It's best to use ordered aesthetics (for example `size` or `alpha`, i.e. transparency) for continuous variables and unordered aesthetics (for example `shape`) for categorical variables. `color` can be ordered or not ordered depending on what color scheme you are using (the same goes for `fill`, which we will talk about later.) 

```{r}
# This is a not-ideal use of size:
ggplot(data = iris) +
  geom_point(mapping = aes(x = Sepal.Length, y = Sepal.Width, size = Species))
# ggplot2 helpfully warns you that "Using size for a discrete variable is not advised."

# Here is a better use of size, mapped to a continuous variable:
ggplot(data = iris) +
  geom_point(mapping = aes(x = Sepal.Length, y = Sepal.Width, size = Petal.Width))

# Here color is used for a continuous variable:
ggplot(data = iris) +
  geom_point(mapping = aes(x = Sepal.Length, y = Sepal.Width, color = Petal.Width))
```

You can specify as many aesthetics as you like, although more than three or four tends to get very confusing...

```{r}
# This is maybe okay:
ggplot(data = iris) +
  geom_point(mapping = aes(x = Sepal.Length, y = Sepal.Width, color = Species, size = Petal.Width))

# This is probably too much for the person looking at your plot to take in all at once.
ggplot(data = iris) +
  geom_point(mapping = aes(x = Sepal.Length, y = Sepal.Width, color = Species, size = Petal.Width, alpha = Petal.Length))
```

Part of making good plots is figuring out how much information you can present in a single plot while still keeping things readable and easy to understand.

Getting back to the nuts and bolts of aesthetics and mapping, you can also manually set an aesthetic without mapping it to a variable - for instance, if you wanted to make all of the points on your plot blue. To manually set an aesthetic, provide the name of the aesthetic as a separate argument (*not* part of the `mapping = aes()` stuff) and then the value you want that aesthetic to take. 

```{r}
# set the color of the points
ggplot(data = iris) +
  geom_point(mapping = aes(x = Sepal.Length, y = Sepal.Width), color = "blue")

# set the transparency of the points
ggplot(data = iris) +
  geom_point(mapping = aes(x = Sepal.Length, y = Sepal.Width), alpha = 0.5)
# some of the points in this plot look different because in some places multiple semi-transparent points are stacked up on top of each other.

# set the shape of the points
ggplot(data = iris) +
  geom_point(mapping = aes(x = Sepal.Length, y = Sepal.Width), shape = 2)

# set the size of the points
ggplot(data = iris) +
  geom_point(mapping = aes(x = Sepal.Length, y = Sepal.Width), size = 3)

# you can even manually set x and/or y, though this isn't very helpful in a scatterplot
ggplot(data = iris) +
  geom_point(mapping = aes(x = Sepal.Length), y = 1)
```

These examples just demonstrate how to use the `mapping` argument and `aes()` function, and how to set aesthetics manually. The particular set of aesthetics available to you depends on what geom function you are using. We'll talk about geoms next.

## 2. Geoms

We've already used one geom function, `geom_point()`, to make a scatterplot. Other geoms can be used to make lots of other kinds of plots; here are a few examples. Take a look at the `ggplot2` cheatsheet (https://github.com/rstudio/cheatsheets/blob/main/data-visualization-2.1.pdf) for even more geoms and to see what aesthetics are available for each geom.

### Histograms

```{r}
ggplot(data = iris) +
  geom_histogram(mapping = aes(x = Sepal.Length))
# `geom_histogram()` defaults to making 30 bins of equal size, but you can control this with the `binwidth` argument

ggplot(data = iris) +
  geom_histogram(mapping = aes(x = Sepal.Length), binwidth = 0.1)

# `color` now controls the border around each bar
ggplot(data = iris) +
  geom_histogram(mapping = aes(x = Sepal.Length, color = Species), binwidth = 0.1)

# `fill` controls the shading of the center of the bar
ggplot(data = iris) +
  geom_histogram(mapping = aes(x = Sepal.Length, fill = Species), binwidth = 0.1)

# both `color` and `fill` will result in stacked bars
```

### Bar plots

Where histograms take a continuous variable on the x axis and divide it into bins in order to count the number of observations in each bin, bar plots take a categorical variable on the x axis and count up the number of observations in each category.

```{r}
# We will use the mtcars data for this
data(mtcars)
head(mtcars)
# Some of these numeric columns actually contain categorical data, so let's make them factors
mtcars$am <- as.factor(mtcars$am)
mtcars$vs <- as.factor(mtcars$vs)

ggplot(data = mtcars) +
  geom_bar(mapping = aes(x = am))

# if you have multiple categorical variables, `color` and `fill` will result in stacked bars
ggplot(data = mtcars) +
  geom_bar(mapping = aes(x = am, fill = vs))
```

### Boxplots

```{r}
ggplot(data = iris) +
  geom_boxplot(mapping = aes(x = Species, y = Sepal.Length))

# if you have multiple categorical variables, `color` and `fill` will result in multiple side-by-side boxplots for each category
ggplot(data = mtcars) +
  geom_boxplot(mapping = aes(x = am, y = mpg, color = vs))
ggplot(data = mtcars) +
  geom_boxplot(mapping = aes(x = am, y = mpg, fill = vs))
```

### Lines

`ggplot2` can draw lines in lots of different ways. We'll look at two main ways: connecting points and smoothing.

First, you can connect a series of observations with `geom_path()`:

```{r}
data(airquality)
filter(airquality, Month == 5) %>%
  ggplot() +
  geom_path(mapping = aes(x = Day, y = Temp))
# By the way, yes, you can pipe data from another function into ggplot. 
# After all, it's a tidyverse function - the first argument is data!
```

Second, you can fit a curve through a cloud of observations with `geom_smooth()`:

```{r}
ggplot(data = iris) +
  geom_smooth(mapping = aes(x = Sepal.Length, y = Sepal.Width))

# geom_smooth has many different smoothing methods, "loess" is the default but often it will be more useful to find a line of best fit with method = "lm"
ggplot(data = iris) +
  geom_smooth(mapping = aes(x = Sepal.Length, y = Sepal.Width), method = "lm")

# you can also map the `color` aesthetic to a categorical variable in order to fit a separate line/curve to each category
ggplot(data = iris) +
  geom_smooth(mapping = aes(x = Sepal.Length, y = Sepal.Width, color = Species), method = "lm")
```

### Combining multiple geoms

The great thing about the layered grammar of graphics is that you can use multiple geoms to create different layers in a single plot. The linear models we made above aren't very meaningful unless we can also see the points, so let's use `geom_point()` and `geom_smooth()` together to make a better plot:

```{r}
ggplot(data = iris) +
  geom_point(mapping = aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_smooth(mapping = aes(x = Sepal.Length, y = Sepal.Width, color = Species), method = "lm")
```

If you're using multiple geoms together and they have aesthetics in common, like `x`, `y`, and `color` in the example above, you can also put these in the call to `ggplot()` and they will be inherited by subsequent layers of the plot:

```{r}
ggplot(data = iris, mapping = aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point() +
  geom_smooth(method = "lm")
```

And the opposite also works - you can specify `data` separately for each geom. (This is especially useful if you're plotting data from multiple data frames in a single plot.)

```{r}
ggplot() +
  geom_point(data = iris, mapping = aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_smooth(data = iris, mapping = aes(x = Sepal.Length, y = Sepal.Width, color = Species), method = "lm")
```

### Making Brissonlab-style mean bars

Points and a mean bar are the customary way to present data in the Brisson lab. `ggplot2` doesn't have a built-in function that does this, but I found a way to jury-rig it using `geom_point()` and some special arguments to `geom_crossbar()`:

```{r}
ggplot(data = mtcars, mapping = aes(x = am, y = mpg)) +
  geom_point() +
  geom_crossbar(stat = "summary", fun = mean)
```


## 3. The `position` argument

The `position` argument controls where different graphical objects in a single layer fall out relative to each other. It behaves somewhat differently for different geoms, so we'll run through a few examples.

For scatterplots:

```{r}
# the default is "identity", but as we saw before this can result in points stacked on top of each other
ggplot(data = iris) +
  geom_point(mapping = aes(x = Sepal.Length, y = Sepal.Width), alpha = 0.5)

# position = "jitter" adds a little bit of random noise to each point so they don't end up on top of each other
ggplot(data = iris) +
  geom_point(mapping = aes(x = Sepal.Length, y = Sepal.Width), alpha = 0.5, position = "jitter")

# you can create this same effect using geom_jitter()
ggplot(data = iris) +
  geom_jitter(mapping = aes(x = Sepal.Length, y = Sepal.Width), alpha = 0.5)
```

For histograms:

```{r}
# the default is stacked bars
ggplot(data = iris) +
  geom_histogram(mapping = aes(x = Sepal.Length, fill = Species), binwidth = 0.1)

# position = "dodge" can be useful for placing bars side-by-side
ggplot(data = iris) +
  geom_histogram(mapping = aes(x = Sepal.Length, fill = Species), binwidth = 0.1, position = "dodge")

# position = "identity" can work if you make the bars transparent - it overlaps the bars
ggplot(data = iris) +
  geom_histogram(mapping = aes(x = Sepal.Length, fill = Species), alpha = 0.5, binwidth = 0.1, position = "identity")
```

For bar plots:

```{r}
# the default is stacked bars
ggplot(data = mtcars) +
  geom_bar(mapping = aes(x = am, fill = vs))

# position = "dodge" puts the bars next to each other instead
ggplot(data = mtcars) +
  geom_bar(mapping = aes(x = am, fill = vs), position = "dodge")

# position = "fill" leaves the bars stacked but converts the count to a ratio out of 1.0 so that each stack of bars is the same height
ggplot(data = mtcars) +
  geom_bar(mapping = aes(x = am, fill = vs), position = "fill")

# position = "identity" overlaps the bars; this can work if you make them transparent
ggplot(data = mtcars) +
  geom_bar(mapping = aes(x = am, fill = vs), alpha = 0.5, position = "identity")
```

For boxplots:

```{r}
# default is side-by-side
ggplot(data = mtcars) +
  geom_boxplot(mapping = aes(x = am, y = mpg, fill = vs))

# position = "identity" overlaps the boxplots, but this can also work if you make them transparent... I guess. I don't find this one particularly helpful.
ggplot(data = mtcars) +
  geom_boxplot(mapping = aes(x = am, y = mpg, fill = vs), alpha = 0.5, position = "identity")
```


## 4. The `stat` argument

The `stat` argument controls whether a statistic is applied to the data when making a layer. In general, most geoms have a good default `stat` - for example, `geom_histogram()` applies the `count()` stat in order to get a count of how many observations are in each bin. `geom_bar()` does the same for each category. But sometimes you might want to be able to turn that off, for example if you are making a bar plot and already have a variable that you want the height of the bars to represent:

```{r}
mtcars <- rownames_to_column(mtcars, "model")

ggplot(data = mtcars) +
  geom_bar(mapping = aes(x = model, y = mpg), stat = "identity")
```

We used this earlier to make `geom_crossbar()` plot means; in that case, we used `stat = "summary"`.

## 5. Faceting

Changing directions now, let's talk about faceting. Plots can get pretty messy if you have more than three aesthetics (for example, x, y, and color). Faceting is a way to display more variables at the same time while still keeping your plot readable. It can also be useful if you want to have a series of plots next to each other, each corresponding to one level of a factor.

```{r}
ggplot(data = airquality) +
  geom_path(mapping = aes(x = Day, y = Temp)) +
  facet_grid(.~Month)
```

You can facet by one or by two variables - take a look at the `ggplot2` cheatsheet for more information.

## 6. Coordinate systems

This isn't something you're likely to use very often; the default coordinate system is Cartesian coordinates, which work great for almost everything. 

However, `coord_flip()` can be useful to make sideways bar plots or boxplots:

```{r}
ggplot(data = mtcars) +
  geom_bar(mapping = aes(x = model, y = mpg), stat = "identity") +
  coord_flip()

ggplot(data = iris) +
  geom_boxplot(mapping = aes(x = Species, y = Sepal.Length)) +
  coord_flip()
```

And `coord_polar()` can be useful on the off chance that you need to make a pie chart or coxcomb chart:

```{r}
# pie chart
ggplot(data = mtcars) +
  geom_bar(mapping = aes(x = "", fill = am), position = "fill", width = 1) +
  coord_polar("y")

# coxcomb chart
mtcars$carb <- as.factor(mtcars$carb)
ggplot(data = mtcars) +
  geom_bar(mapping = aes(x = carb, fill = carb), position = "dodge", width = 1) +
  coord_polar()

# warning: ggplot2 tends to get a little confused about how to label these kinds of plots, and I will freely admit I'm not an expert at controlling its behavior here. 
```

# PART 3: Making your plots pretty and readable

`ggplot2`'s grammar of graphics makes it possible for you to create any sort of plot you can imagine... but once you have all of the elements on the page, there are still lots of things you can tweak to make your plots *pretty*. We'll go over a couple major categories: choosing colors, controlling axis labels/ranges/breakpoints, annotating a plot with text/shapes, dealing with legends, and using themes to control the "style" of the plot.

## 1. Color

By default, `ggplot2` uses colors that are evenly distributed around the color wheel for categorical variables, and a range of dark to light blues for continuous variables. While these defaults are fine for just getting a first glance at your data, you'll probably want to choose your own colors when making a plot for someone else to look at. 

If you have `color` and/or `fill` mapped to a categorical variable, use the functions `scale_color_manual()` and/or `scale_fill_manual()` to choose your own colors:

```{r}
# specify the 3 colors to use for the 3 species
ggplot(data = iris) +
  geom_histogram(mapping = aes(x = Sepal.Length, fill = Species), binwidth = 0.1) +
  scale_fill_manual(values = c("darkgreen", "forestgreen", "chartreuse3"))
```

Note that color names are supplied to the `values` argument as a vector of strings. For the names of colors in R, I highly recommend this resource: http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf

You can also use hex color codes, if you're familiar with those.

```{r}
ggplot(data = iris) +
  geom_histogram(mapping = aes(x = Sepal.Length, fill = Species), binwidth = 0.1) +
  scale_fill_manual(values = c("#228B22", "#32CD32", "#00FF00"))
```

If you have `color` and/or `fill` mapped to a continuous variable, use the `scale_color_gradient()` / `scale_fill_gradient()` family of functions.

```{r}
# to get a gradient between two specified colors
ggplot(data = iris) +
  geom_point(mapping = aes(x = Sepal.Length, y = Sepal.Width, color = Petal.Width)) +
  scale_color_gradient(low = "mediumpurple1", high = "purple4")

# use scale_color_gradient2(low = , mid = , high = ) for a diverging color gradient centered on zero
```

Alternately, use the `viridis` scales for beautiful and colorblindness-friendly colors.

```{r}
# scale_color_viridis_c or scale_fill_viridis_c for continuous scales
ggplot(data = iris) +
  geom_point(mapping = aes(x = Sepal.Length, y = Sepal.Width, color = Petal.Width)) +
  scale_color_viridis_c()

# scale_color_viridis_d or scale_fill_viridis_d for discrete scales
ggplot(data = iris) +
  geom_histogram(mapping = aes(x = Sepal.Length, fill = Species), binwidth = 0.1) +
  scale_fill_viridis_d()

# there are also alternative viridis scales named magma, plasma, and inferno
# use the `option` argument to select one of these
ggplot(data = iris) +
  geom_point(mapping = aes(x = Sepal.Length, y = Sepal.Width, color = Petal.Width)) +
  scale_color_viridis_c(option = "plasma")
```

For more information on the `viridis` scales, see: https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html

## 2. Axes

`ggplot2` automatically labels axes and legends with the name of the variable that is mapped to each aesthetic, but your variable names might not be meaningful to another person. To re-label your axes, use `labs()`:

```{r}
ggplot(data = mtcars) +
  geom_bar(mapping = aes(x = am, fill = vs)) +
  labs(x = "Transmission type", fill = "Engine \nconfiguration")
# \n here creates a line break between "Engine" and "configuration" so things fit a little more comfortably on the page
```

Similarly, `ggplot2` automatically labels the bins for a categorical variable with the values of that variable. To change this, you can use the `labels` argument of the `scale_x_discrete()` function. A similar method can be used to change the labels associated with `color` or `fill`. (Use `scale_fill_hue()` if you want to keep the default `ggplot2` colors; otherwise, use `scale_fill_manual()` and change the labels at the same time as you change the colors.)

```{r}
ggplot(data = mtcars) +
  geom_bar(mapping = aes(x = am, fill = vs)) +
  labs(x = "Transmission type", fill = "Engine \nconfiguration") +
  scale_x_discrete(labels = c("automatic", "manual")) +
  scale_fill_hue(labels = c("V", "standard"))
```

It's also possible to rotate x-axis labels using `theme(axis.text.x = )` (and similar for the y-axis).

```{r}
ggplot(data = mtcars) +
  geom_bar(mapping = aes(x = model, y = mpg), stat = "identity") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

For continuous scales, it can be useful to control the range and breakpoints. Range is fairly self-explanatory; we will demonstrate how it can be set with `scale_y_continuous()`.

```{r}
# make the y-axis on this plot go up to 20, instead of ending immediately above the data
ggplot(data = mtcars) +
  geom_bar(mapping = aes(x = am, fill = vs)) +
  labs(x = "Transmission type", fill = "Engine \nconfiguration") +
  scale_x_discrete(labels = c("automatic", "manual")) +
  scale_y_continuous(limits = c(0,20))
# ggplot2 tends to be good at scaling the axes to fit your data, so this is often not terribly useful on its own.
# However, it can be extremely helpful when working with percentage data (if you want to show the full 0%-100% range), or when you plan to put two plots side-by-side and want their scales to match.
```

`scale_y_continuous()` and the like can also be used to control breakpoints. `ggplot2` recognizes both major and minor breakpoints. In the default theme, major breaks are indicated by a labeled tickmark and a thick white gridline; minor breaks receive a thinner gridline and no tickmark.

```{r}
ggplot(data = mtcars) +
  geom_bar(mapping = aes(x = vs, fill = cyl)) +
  labs(x = "Engine configuration", fill = "Number of \ncylinders") +
  scale_x_discrete(labels = c("V", "standard")) +
  scale_y_continuous(limits = c(0,20), breaks = seq(0,20, by = 10), minor_breaks = seq(0, 20, by = 1))
# I am using seq() to create a sequence of numbers that denote break locations, but you can also supply any vector of numbers
```

There are so many little tiny ways in which axis labels can be modified, so it's a good idea to read the `ggplot2` cheatsheet and documentation and spend some time googling if there's something particular you would like to do. Chances are, if you can imagine it, there's a way to do it with `ggplot2`.

## 3. Annotations

Another thing that can be useful to add to a plot is annotations. These are marks (usually text, lines, or shapes) that you manually create on the plot rather than getting their position from the data.

### Text annotations

To add text annotations, use `annotate("text")`.

```{r}
ggplot(data = iris) +
  geom_point(mapping = aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  annotate("text", x = 4.8, y = 4.25, label = "setosa") +
  annotate("text", x = 7.25, y = 2.25, label = "versicolor and \nvirginica")
# "text" tells it that you are creating a text annotation
# x and y tell it the coordinates the annotation should be placed at
# label tells it what the text annotation should say
```

### Lines and shapes

To make a line or rectangle on a plot, you can continue to use `annotate()`:

```{r}
# random horizontal line
ggplot(data = iris) +
  geom_point(mapping = aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  annotate("segment", x = 4.5, xend = 7.5, y = 3, yend = 3)

# random vertical line
ggplot(data = iris) +
  geom_point(mapping = aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  annotate("segment", x = 6, xend = 6, y = 2, yend = 4.5)

# random diagonal line
ggplot(data = iris) +
  geom_point(mapping = aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  annotate("segment", x = 4.5, xend = 7, y = 2.5, yend = 4.5)

# random rectangle
ggplot(data = iris) +
  geom_point(mapping = aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  annotate("rect", xmin = 6, xmax = 8, ymin = 4, ymax = 4.5)
# for a rectangle, use xmin/xmax/ymin/ymax instead of x/xend/y/yend

# can control color, fill, etc. of annotations
ggplot(data = iris) +
  geom_point(mapping = aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  annotate("rect", xmin = 6, xmax = 8, ymin = 4, ymax = 4.5, color = "black", fill = NA)
```

There are also specialized functions you can use for making horizontal lines (`geom_hline()`), vertical lines (`geom_vline()`), and rectangles (`geom_rect()`). 

```{r}
# horizontal line
ggplot(data = iris) +
  geom_point(mapping = aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_hline(yintercept = 3)
# note that this horizontal line just goes the full width of the plot

# vertical line
ggplot(data = iris) +
  geom_point(mapping = aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_vline(xintercept = 6)
# similarly, this vertical line goes the full height of the plot

# rectangle
ggplot(data = iris) +
  geom_point(mapping = aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_rect(xmin = 6, xmax = 8, ymin = 4, ymax = 4.5)
```

## 4. Legends

`ggplot2` automatically creates legends for you when you use non-xy aesthetics. By default, the legend is positioned to the right of the plot, but this can be adjusted manually:

```{r}
ggplot(data = iris) +
  geom_histogram(mapping = aes(x = Sepal.Length, fill = Species), binwidth = 0.1) +
  theme(legend.position = "left")
# and likewise for top and bottom
```

And you can use the same method to get rid of the legend entirely, if you like:

```{r}
ggplot(data = iris) +
  geom_histogram(mapping = aes(x = Sepal.Length, fill = Species), binwidth = 0.1) +
  theme(legend.position = "none")
```

This is equivalent to adding `show.legend = FALSE` to any given geom. Using `theme(legend.position = "none")` to get rid of the legend will remove the legend for the entire plot, while `show.legend = FALSE` will just prevent that particular geom from contributing a legend.

```{r}
# This plot has two geoms contributing to the legend: geom_point and geom_smooth
ggplot(data = iris, mapping = aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point() +
  geom_smooth(method = "lm")

# Get rid of just the geom_point legend
ggplot(data = iris, mapping = aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point(show.legend = FALSE) +
  geom_smooth(method = "lm")

# Get rid of just the geom_smooth legend
ggplot(data = iris, mapping = aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point() +
  geom_smooth(method = "lm", show.legend = FALSE)

# Get rid of both legends separately
ggplot(data = iris, mapping = aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point(show.legend = FALSE) +
  geom_smooth(method = "lm", show.legend = FALSE)

# Get rid of all legends in one command with theme(legend.position = "none")
ggplot(data = iris, mapping = aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme(legend.position = "none")
```

## 5. Themes

`ggplot2` provides a number of "themes" which can be used to control the overall style of a plot. Try running the following to see what these different built-in themes look like. (For even more themes, take a look at the `ggthemes` package!)

```{r}
# theme_gray() is the default ggplot2 theme
ggplot(data = iris) +
  geom_point(mapping = aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  theme_gray()

ggplot(data = iris) +
  geom_point(mapping = aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  theme_classic()

ggplot(data = iris) +
  geom_point(mapping = aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  theme_bw()

ggplot(data = iris) +
  geom_point(mapping = aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  theme_dark()

ggplot(data = iris) +
  geom_point(mapping = aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  theme_void()
```

You can also manually control pretty much every aspect of a plot's appearance and style using the `theme()` function. We previously used `theme()` to rotate axis text labels and to control the position of the legend. To see all of the other arguments this function can take, try using `?theme` to pull up the documentation:

```{r}
?theme
```

For many `theme()` arguments, you will need to use a helper `element_` function: `element_rect()`, `element_line()`, `element_text()`, etc. To get rid of a particular element, use `element_blank()`. 

```{r}
?element_rect
?element_line
?element_text

# Here are some examples that just tweak one thing at a time so you can see what each bit is doing.

# Starting with this plot:
ggplot(data = iris) +
  geom_point(mapping = aes(x = Sepal.Length, y = Sepal.Width, color = Species))

# panel background
ggplot(data = iris) +
  geom_point(mapping = aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  theme(panel.background = element_rect(fill = "white"))

# axis lines (example with x, y is analogous; use axis.line to control both simultaneously)
ggplot(data = iris) +
  geom_point(mapping = aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  theme(axis.line.x = element_line(color = "black", size = 0.3))

# size of axis labels (example with x, y is analogous; use axis.title to control both simultaneously)
ggplot(data = iris) +
  geom_point(mapping = aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  theme(axis.title.x = element_text(size=20))

# gridlines (example with minor gridlines, major is analogous; use panel.grid to control both simultaneously)
ggplot(data = iris) +
  geom_point(mapping = aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  theme(panel.grid.minor = element_blank())

# Putting it all together:
ggplot(data = iris) +
  geom_point(mapping = aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  theme(panel.background = element_rect(fill = "white"), axis.line = element_line(color = "black", size = 0.3), axis.title = element_text(size=12), panel.grid = element_blank())
```

If you are manually controlling elements of the theme and want to make a set of figures that all go together and have the same theme, it can be simpler to just save the theme to a variable, then add the variable to each plot you make. (This is something Nancy Chen does a lot.)

```{r}
plottheme <- theme(panel.background = element_rect(fill = "white"), axis.line = element_line(color = "black", size = 0.3), axis.title = element_text(size=12), panel.grid = element_blank())

ggplot(data = iris) +
  geom_point(mapping = aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  plottheme

ggplot(data = iris) +
  geom_histogram(mapping = aes(x = Sepal.Length, fill = Species), binwidth = 0.1) +
  plottheme
```

# PART 4: Exporting publication-ready, multipart figures

When you're ready to make figures for a manuscript, you'll often want to combine several plots together into a multipart figure. The `cowplot` package is great for this.

```{r}
#install.packages("cowplot")
library(cowplot)
```

Just as a sidenote - `cowplot` has its own `ggplot2` theme, `theme_cowplot`, and older versions of the `cowplot` package (<1.0.0) will automatically apply this theme to all ggplots when the `cowplot` package is loaded (whether or not you're using a `cowplot` function). However, you can still use `theme()` to override this default and manually control all of the stylistic features of your plots. If you are using a more recent version of the `cowplot` package, this won't be an issue for you.

Now, let's use the `plot_grid` function from the `cowplot` package to make a simple multipart figure.

```{r}
# first save each of the plots you want to include to a variable
a <- ggplot(data = iris) +
  geom_point(mapping = aes(x = Sepal.Length, y = Sepal.Width, color = Species))
b <- ggplot(data = iris) +
  geom_histogram(mapping = aes(x = Sepal.Length, fill = Species), binwidth = 0.1)

# then call plot_grid to combine the plots
plot_grid(a, b, ncol = 2)
# plot_grid can arrange plots in columns, rows, or both.
plot_grid(a, b, nrow = 2)
```

If you want to combine a ggplot with a base R graphic (yes, this has happened to me) you can use `recordPlot()` to save the base R plot to a variable, then use `plot_grid` as above. Alternately, you can write a function that will make your base R plot, then refer to it in your call to `plot_grid` in the same way you would refer to a plot saved as variable. Either way, you will need to install and load package `gridGraphics` in order for `plot_grid` to handle base R graphics.

```{r}
#install.packages("gridGraphics")
library(gridGraphics)

par(mar = c(6,10,1,1))
plot(x=iris$Sepal.Length, y=iris$Sepal.Width)
base_R_plot <- recordPlot()
plot_grid(a, base_R_plot, ncol = 2)

# Using a function to make the base R plot seems to work better a lot of the time.
base_R_plot_2 <- function() {
  par(mar = c(6,10,1,1))
  plot(x=iris$Sepal.Length, y=iris$Sepal.Width)
}
plot_grid(a, base_R_plot_2, ncol = 2)
```

In any case, for a manuscript, you might want to label the parts of your figure (for example, A and B).

```{r}
# labels = "AUTO" will automatically label the parts of your figure with capital letters
plot_grid(a, b, ncol = 2, labels = "AUTO")

# labels = "auto" will do the same but with lowercase
plot_grid(a, b, ncol = 2, labels = "auto")

# or you can manually create whatever labels you like
plot_grid(a, b, ncol = 2, labels = c("A", "B"))
```

This looks a bit silly; since the same color scheme is used in both plots, we don't really need both legends! We can remove the legend from one plot using `theme(legend.position = "none")`, as we saw above. Just add this to the existing plot to produce a legendless version.

```{r}
a_nolegend <- a + theme(legend.position = "none")

plot_grid(a_nolegend, b, ncol = 2, labels = c("A", "B"))
```

Another way to deal with the legend is to treat it as a separate object. You can "capture" the legend off of one of your plots using the `get_legend()` function, then plot your captured legend alongside legendless versions of both plots:

```{r}
ab_legend <- get_legend(b)

b_nolegend <- b + theme(legend.position = "none")

plot_grid(a_nolegend, b_nolegend, ab_legend, ncol = 3, labels = c("A", "B", ""))
```

Notice that we now have 3 columns, with `a`, `b`, and the legend each occupying their own column. We've labeled the two plots "A" and "B" and supplied a blank label for the legend.

By default, `plot_grid()` makes all of the columns in a plot equal width (and all of the rows in a plot equal height), but you might want some of your columns to be wider than others. In this case, the legend probably doesn't need to be as wide as the plots. We can use `rel_widths` (or, if the plots are arranged in columns, `rel_heights`) to allow more space for the two plots and less space for the legend:

```{r}
plot_grid(a_nolegend, b_nolegend, ab_legend, ncol = 3, labels = c("A", "B", ""), rel_widths = c(1, 1, 0.5))
```

When you're satisfied with your plot, export it using `ggsave()`, `cowplot`'s `save_plot()` function, or `pdf()`. Be aware that `ggsave()` and `save_plot()` only work on ggplots and plots created with `plot_grid` - if you're also exporting plain base R graphics, it might be better to use `pdf()` to export everything so you don't have to switch back and forth. `pdf()` will always create PDF files, while `ggsave()` and `save_plot()` will guess the file type based on what extension you give the filename.

```{r}
# ggsave defaults to saving the last plot you made with the dimensions of your current plotting window
ggsave("ggsave_test.pdf")
# can specify width and height with ggsave("ggsave_test.pdf", width = , height = , units = "in")

# for save_plot(), need to save the multipart plot to a variable first
ab <- plot_grid(a, b, ncol = 2, labels = c("A", "B"), rel_widths = c(1, 1.65))
# save_plot also doesn't have as useful of defaults - you will need to at least specify an aspect ratio, and can optionally specify base_width and base_height
save_plot("save_plot_test.pdf", ab, base_aspect_ratio = 2)

# pdf() method is a little more involved...
# first call pdf() and tell it what to name the file and the dimensions
pdf("pdf_test.pdf", width = 8, height = 4)
# run the plotting command
plot_grid(a, b, ncol = 2, labels = c("A", "B"), rel_widths = c(1, 1.65))
# then do dev.off() to terminate the graphical device
dev.off()
```


# Resources

`ggplot2` cheatsheet: https://github.com/rstudio/cheatsheets/blob/main/data-visualization-2.1.pdf

`ggplot2` function reference page: https://ggplot2.tidyverse.org/reference/

List of colors in R: http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf

Chapters 3 and 28 of the book *R for Data Science* (Wickham and Grolemund 2017), which give tons more examples of how to use `ggplot2`: https://r4ds.had.co.nz/data-visualisation.html and https://r4ds.had.co.nz/graphics-for-communication.html

Intro to `viridis`: https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html

More information on the `cowplot` package: https://cran.r-project.org/web/packages/cowplot/vignettes/introduction.html


# References

Wickham, H. 2010. A Layered Grammar of Graphics. Journal of Computational and Graphical Statistics **19**:3-28

Wickham, H., and G. Grolemund. 2017. R for Data Science: Import, Tidy, Transform, Visualize, and Model Data. O'Reilly Media, Inc.

